---
title: "Lesson 15: Working with categorical variables, dummy coding, ANOVA, Regression and the GLM"
author: "Melinda K. Higgins, PhD."
date: "October 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
# set options
knitr::opts_chunk$set(echo = TRUE)

# load packages needed
library(tidyverse)
library(haven)
library(car)
library(olsrr)
```

## Load data, Recode race

Load the HELP dataset and recode race - merge the "other" and "hispanic" races together since both have low % in this dataset. Make a table to check recoding and check class of variable.

```{r}
helpdat <- haven::read_spss("helpmkh.sav")

# let's merge hispanic/other together for race
helpdat$race3 <- ""
helpdat$race3[helpdat$racegrp == "black"] <- "Black"
helpdat$race3[helpdat$racegrp == "white"] <- "White"
helpdat$race3[helpdat$racegrp == "hispanic"] <- "Hisp/Other"
helpdat$race3[helpdat$racegrp == "other"] <- "Hisp/Other"
```

### Table of `race3`

```{r}
table(helpdat$race3)
class(helpdat$race3)
```

Let's also creat a numeric coded variable and check the new variable & from the numeric variable also make another variable that is a `factor` type class variable.

```{r}
helpdat$race3num[helpdat$race3 == "Black"] <- 1
helpdat$race3num[helpdat$race3 == "White"] <- 2
helpdat$race3num[helpdat$race3 == "Hisp/Other"] <- 3

helpdat$race3numF <- factor(helpdat$race3num,
                            levels=c(1,2,3),
                            labels=c("1. Black",
                                     "2. White",
                                     "3. Hisp/Other"))
```

### Table of `race3num`

```{r}
table(helpdat$race3num)
class(helpdat$race3num)
```

### Table of `race3numF`

```{r}
table(helpdat$race3numF)
class(helpdat$race3numF)
```

Get a summary table of the means and SD for sexrisk by `race3`, `racenum` or `race3numF` - to do this we will use the `dplyr` functions `group_by()` and `summarise()`.

### Summary stats for `sexrisk` by `race3`

```{r}
helpdat %>%
  select(sexrisk, race3) %>%
  group_by(race3) %>%
  summarise(mean=mean(sexrisk), 
            sd=sd(sexrisk))
```

### Summary stats for `sexrisk` by `race3num`

```{r}
helpdat %>%
  select(sexrisk, race3num) %>%
  group_by(race3num) %>%
  summarise(mean=mean(sexrisk), 
            sd=sd(sexrisk))
```

### Summary stats for `sexrisk` by `race3numF`

```{r}
helpdat %>%
  select(sexrisk, race3numF) %>%
  group_by(race3numF) %>%
  summarise(mean=mean(sexrisk), 
            sd=sd(sexrisk))
```

Run an ANOVA for `sexrisk` compare for the different variables `race3`, `race3num` and `race3numF`

### ANOVA for `sexrisk` by `race3` - character variable

```{r}
# race3 as a character variable
aov1 <- aov(sexrisk ~ race3, data=helpdat)
aov1
summary(aov1)
model.tables(aov1)
TukeyHSD(aov1)
with(data=helpdat, 
     pairwise.t.test(sexrisk, race3, 
                     p.adj="none", paired=FALSE))
car::Anova(aov1, type=3)
```

### ANOVA for `sexrisk` by `race3num` - numeric variable

```{r}
# race3num as a numeric variable
aov2 <- aov(sexrisk ~ race3num, data=helpdat)
aov2
summary(aov2)
model.tables(aov2) # odd results
#TukeyHSD(aov2) # not run
with(data=helpdat, 
     pairwise.t.test(sexrisk, race3num, 
                     p.adj="none", paired=FALSE)) # this works
car::Anova(aov2, type=3)
```

### ANOVA for `sexrisk` by `race3numF` - numeric variable

```{r}
# race3numF as a Factor type/class variable
aov3 <- aov(sexrisk ~ race3numF, data=helpdat)
aov3
summary(aov3)
model.tables(aov3)
TukeyHSD(aov3)
with(data=helpdat, 
     pairwise.t.test(sexrisk, race3numF, 
                     p.adj="none", paired=FALSE))
car::Anova(aov3)
```

### Create dummy variables for `race3`

```{r}
# add dummy variables
helpdat <- helpdat %>%
  mutate(race_black = race3num==1,
         race_white = race3num==2,
         race_otherhisp = race3num==3)

table(helpdat$race3num)
table(helpdat$race_black)
table(helpdat$race_white)
table(helpdat$race_otherhisp)
```

### Regression approach using dummy variables

Set `white` as the reference `race3` category - so use the other 2 dummy vars

```{r}
lm1 <- lm(sexrisk ~ race_black + race_otherhisp, data=helpdat)
lm1
summary(lm1)
```

Set `otherhisp` as the reference `race3` category - so use the other 2 dummy vars

```{r}
lm2 <- lm(sexrisk ~ race_black + race_white, data=helpdat)
lm2
summary(lm2)
```

### Regression approach with a `factor` type variable

In R when you put a **FACTOR** type variable into a model, the dummy coding happens on the fly this also works if the variable is **character** type so R recognizes the categories and treats them as a `factor` for modeling purposes. Notice that the 1st category for `black` is used as the reference category by default.

Regression approach using `race3numF` which is a `factor` type variable

```{r}
lm3f <- lm(sexrisk ~ race3numF, data=helpdat)
lm3f
summary(lm3f)
car::Anova(lm3f, type=3)
```

Regression approach using `race3` which is a `character` type variable

```{r}
lm3c <- lm(sexrisk ~ race3, data=helpdat)
lm3c
summary(lm3c)
car::Anova(lm3c, type=3)
```

Regression approach using `race3num` which is a `numeric` type variable. When the variable is numeric and is not treated as a factor, you get a single coefficient and 1 slope "effect".

```{r}
lm4 <- lm(sexrisk ~ race3num, data=helpdat)
lm4
summary(lm4)
car::Anova(lm4, type=3)
```

Use the `ols_regress()` function from the `olsrr` package to get more detailed outputs for comparison if you wish.

### Regression results for `lm1` using dummy variable coding

```{r}
olsrr::ols_regress(lm1)
```

### Regression results for `lm3f` using `race3numF` as a `factor` variable

```{r}
olsrr::ols_regress(lm3f)
```

### Regression results for `lm4` using `race3num` as a `character` variable

```{r}
olsrr::ols_regress(lm4)
```

